---
title: "EPAC adults and climate indices"
format: html
---

# Setup

loading packages:

```{r}
library(here)
library(readxl)
library(tidyverse)
library(ggplot2)
library(viridis)
library(leaflet)
library(broom.mixed)
library(DT)
library(mgcv)
library(purrr)
library(patchwork)
library(ggeffects)
library(gratia)
library(janitor)
library(sf)
library(plotly)
```


loading data:

```{r}
raw_data <- read_excel(here("data", "raw", "krill_size.xlsx"))
raw_data_2025 <- read_excel(here("data", "raw", "2025_krill_size.xlsx"))
moci_data <- read.csv(here("data", "raw", "CaliforniaMOCI_long.csv"))
loc_data <- read_excel(here("data","raw", "locdata.xlsx"))
climate_raw <- read_excel(here("data","raw", "climate_ocean_indices.xlsx"))
tow_data <- read_excel(here("data","raw", "tow_data.xlsx"))

```

# Creating main dataframe

## Filtering data

Extracting year and month information from the cruise column and converting it to a convenient format. Creating base_date in the yyyy-mm-01 format to make sure it is treated as date in R. Used "cruise" column as reference rather than "date", since some sampling is done early in the next month (e.g. sampling date is 7/1 but it's from june cruise)

cleaning up 2025 data:

```{r}
data_2025 <-
clean_names(raw_data_2025) %>%
  rename(jar = id_jar)

  data_2025$line <- as.character(data_2025$line)
```


adding 2025 data:

```{r}
data_all <- bind_rows (raw_data, data_2025) %>%
  mutate(base_date = make_date(
      year  = 2000 + as.numeric(substr(cruise, 4, 5)),
      month = as.numeric(substr(cruise, 6, 7)),
      day   = 1))%>% 
  relocate(base_date, .after = date)
```  

adding time and location data:

```{r}
tow_data <- clean_names(tow_data) %>%
  rename(jar = id_jar)

data_all <- data_all %>%
  left_join(
    tow_data %>% select(lat_in, lon_in, jar, time_in, depth_mean), by = "jar")%>%
  mutate(time_in = hms::as_hms(time_in))
```

adding D/N for times

```{r}
data_all <- data_all %>%
  mutate(hour = hour(time_in),
  diel = case_when(
      hour >= 6 & hour < 18 ~ "day",
      TRUE ~ "night"))
```


adding MOCI

```{r}
# make lagged MOCI table in the same style as df_indices
moci_indices <- moci_data %>%
  select(Year, Season, North.California..38.42N.) %>%
  rename(value_raw = North.California..38.42N.) %>%
  mutate(
    season_order = case_when(
      Season == "JFM" ~ 1,
      Season == "AMJ" ~ 2,
      Season == "JAS" ~ 3,
      Season == "OND" ~ 4
    )
  ) %>%
  arrange(Year, season_order) %>%
  mutate(
    MOCI_lag0 = value_raw,
    MOCI_lag1 = lag(value_raw, 1),
    MOCI_lag2 = lag(value_raw, 2),
    MOCI_lag3 = lag(value_raw, 3)
  ) %>%
  select(Year, Season, starts_with("MOCI_lag")) %>%
  pivot_longer(
    cols = starts_with("MOCI_lag"),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    index = "MOCI",
    lag = as.numeric(gsub("MOCI_lag", "", variable))
  ) %>%
  select(Year, Season, variable, index, lag, value) %>%
  arrange(Year, Season, lag)
```


```{r}
cruise_lookup <- data_all %>%
  mutate(
    Year = year(base_date),
    Month = month(base_date),
    Season = case_when(
      Month %in% c(1, 2, 3) ~ "JFM",
      Month %in% c(4, 5, 6) ~ "AMJ",
      Month %in% c(7, 8, 9) ~ "JAS",
      Month %in% c(10, 11, 12) ~ "OND"
    )
  ) %>%
  select(cruise, Year, Season) %>%
  distinct() 
```

```{r}
moci_indices_with_cruise <- moci_indices %>%
  left_join(cruise_lookup, by = c("Year", "Season")) %>%
  select(-Year, -Season) %>% drop_na()
```
  
organizing the data: cleaning up the names, changing all cruise names to ACC format

```{r}
climate <- climate_raw %>%
  filter(!is.na(cruise)) %>%
  rename_with(~ str_replace(., "-(\\d+)", "_lag\\1")) %>%
    mutate(cruise = case_when(
    cruise == "ACM1705" ~ "ACC1705", 
    cruise == "ACD1807" ~ "ACC1807",
    cruise == "ACM2205" ~ "ACC2205", 
    cruise == "ACD2206" ~ "ACC2206",
    
    TRUE ~ cruise)) 
  
```

adding lags in a separate column

```{r}
df_climate <- climate %>% 
  select(cruise, SOI, SOI_lag1, SOI_lag2, SOI_lag3, PDO, PDO_lag1, PDO_lag2, PDO_lag3, NPGO, NPGO_lag1, NPGO_lag2, NPGO_lag3, CUTI, CUTI_lag1, CUTI_lag2, CUTI_lag3, BEUTI, BEUTI_lag1, BEUTI_lag2, BEUTI_lag3, UI, UI_lag1, UI_lag2, UI_lag3, SEFI_SST_mean, SEFI_SST_anomalies, SEFI_SST_mean_lag1, SEFI_SST_anomalies_lag1, SEFI_SST_mean_lag2, SEFI_SST_anomalies_lag2, SEFI_SST_mean_lag3, SEFI_SST_anomalies_lag3)  %>%
  pivot_longer(
    cols = -cruise,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    index = str_extract(variable, "^[^_]+"),     # PDO, SOI, etc.
    lag = str_extract(variable, "(?<=lag)\\d+")  # 1, 2, 3
  ) %>%
  mutate(
    lag = ifelse(is.na(lag), 0, as.numeric(lag))
  ) %>%

  mutate(index = case_when(
    index == "SEFI" ~ "SST", TRUE ~ index))

```

adding all indices in one df:

```{r}
df_all_indices <- bind_rows(df_climate, moci_indices_with_cruise)
```


Joining all indices with krill data:

```{r}
data_all_climate <- left_join(data_all, df_all_indices, by = "cruise")
```


 
 Filtering months of cruises, only keeping EPAC adults:

```{r}
data_krill <- data_all_climate %>%
  filter(! month(base_date) %in% c(2, 4, 10)) %>%
  mutate(
    type = case_when(
      type == "imm m" ~ "m",
      type == "imm f" ~ "f",
      TRUE ~ type)) %>%
  filter(type %in% c("f", "m")) %>%
  filter(species_name %in% c(
    "EUP Euphausia pacifica"))
```


creating separate datasets for day and night:

```{r}
data_day <- data_krill %>%
  filter(diel == "day")

data_night <- data_krill %>%
  filter(diel == "night")
```

## Size Calculation

Adding columns for width and length midpoint for each size range in each type:

```{r}
df <- data_day%>%
  mutate(length_mid = (length_l + length_h) / 2) %>%
  mutate(width_mid = (width_l + width_h) / 2) %>%
relocate(length_mid, .after = length_h) %>%
  relocate(width_mid, .after = width_h)
```

Calculating volume per each row by considering krill cone-shaped, using width_mid as diameter and length_mid as height:

```{r}
df <- df %>%
  mutate(volume = (1/3) * pi * (width_mid / 2)^2 * length_mid) %>%
  relocate(volume, .after = width_mid) %>%
  filter(!is.na(volume)) 
```

Calculating mean volume per jar: (weighted mean of volume instead of means of l and w first then calculating mean volume with mean l and w)

```{r}
df_vol <- df %>%
  group_by( jar, type, line, cruise, variable, value, index, lag, depth_mean) %>%
  filter(!is.na(tot_num)) %>%
  summarise(n = sum(tot_num), volume_weighted = sum(volume * tot_num) / n,
    .groups = "drop")
```


#Visuals


```{r}
df_vol <- df_vol %>%
  mutate(cruise = factor(cruise, levels = sort(unique(cruise))))
```

```{r}

# One value per cruise for volume_weighted
df_volume <- df_vol %>%
  group_by(cruise, variable, value) %>%
  summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE), .groups = "drop")

p <- plot_ly()

# Fixed weighted volume line
p <- p %>%
  add_trace(
    data = df_volume,
    x = ~cruise,
    y = ~volume_weighted,
    type = "scatter",
    mode = "lines+markers",
    name = "volume_weighted",
    line = list(width = 4)
  )

# Add one hidden trace per variable
vars <- unique(df_volume$variable)

for (v in vars) {
  df_sub <- df_volume %>% filter(variable == v)

  p <- p %>%
    add_trace(
      data = df_sub,
      x = ~cruise,
      y = ~value,
      type = "scatter",
      mode = "lines+markers",
      name = v,
      visible = "legendonly"
    )
}

p <- p %>%
  layout(
    title = "Weighted Volume vs Variables Across Cruises",
    xaxis = list(
      title = "Cruise",
      categoryorder = "array",
      categoryarray = levels(df_vol$cruise)
    ),
    yaxis = list(title = "Value"),
    legend = list(
      itemclick = "toggle",
      itemdoubleclick = "toggleothers"
    )
  )

p
```



# Analysis - other indices



index:


