---
title: "Krill Size"
subtitle: "Changes in krill size in CCE"
author: Lala Movsesian
format: html
---

# Setup

loading packages and data

```{r}
library(here)
library(readxl)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest) 
library(viridis)
library(leaflet)

raw_data <- read_excel(here("data", "raw", "krill_size.xlsx"))
moci_data <- read.csv(here("data", "raw", "CaliforniaMOCI_long.csv"))
loc_data <- read_excel(here("data","raw", "locdata.xlsx"))
```

### Date

Extracting year and month information from the cruise column and converting it to a convenient format. Creating base_date in the yyyy-mm-01 format to make sure it is treated as date in R. Used "cruise" column as reference rather than "date", since some sampling is done early in the next month (e.g. sampling date is 7/1 but it's from june cruise)

```{r}
data_date <- raw_data %>%
  mutate(base_date = make_date(
      year  = 2000 + as.numeric(substr(cruise, 4, 5)),
      month = as.numeric(substr(cruise, 6, 7)),
      day   = 1))%>% 
  relocate(base_date, .after = date)

```

excluding cruises from feb (n=1), apr (n=2), and oct (n=3)

```{r}
df <- data_date %>%
  filter(!month(base_date) %in% c(2, 4, 10))
```

### Life Stage

Changing all imm m to m, and imm f to f for consistency, since some samples include rows labeled as m in lower than 11mm ranges.

```{r}
df <- df %>%
  mutate(
    type = case_when(
      type == "imm m" ~ "m",
      type == "imm f" ~ "f",
      TRUE ~ type))
```

### Species

Filtering data: only including species of interest, only including juv, f and m

```{r}
df <- df %>%
  filter(species_name %in% c(
    "EUP Euphausia pacifica",
    "EUP Nematoscelis difficilis",
    "EUP Thysanoessa spinifera")) %>%
  
  filter(type %in% c("juv", "f", "m"))
```

Adding code names for convenience

```{r}
df <- df %>%
  mutate(
    species_code = case_when(
      species_name == "EUP Euphausia pacifica" ~ "EPAC",
      species_name == "EUP Thysanoessa spinifera" ~ "TSPIN",
      species_name == "EUP Nematoscelis difficilis" ~ "NDIF")) %>%

relocate(species_code, .after = species_name)
```

## MOCI

Joining seasonal MOCI from Northern California with df

```{r}
df <- df %>%
  mutate(
    Year = year(base_date),
    Month = month(base_date),
    Season = case_when(
      Month %in% c(4, 5, 6) ~ "AMJ",
      Month %in% c(7, 8, 9) ~ "JAS"))
    
df <- df %>%
  left_join(
    moci_data %>%
      select(Year, Season, North.California..38.42N.),
    by = c("Year", "Season")) %>%
  rename(moci = North.California..38.42N.) %>%
  rename(season = Season) %>%
  select(-Year, -Month) %>%
  relocate(season, .after = base_date)
```





# Analyses

Testing interannual variability using Linear Mixed Models because:

-   nested data (no need to collapse data to means)

-   not independent data (avoiding pseudoreplication and inflated significance)

-   considering uneven sampling effort

-   more suitable for different months and different numbers of cruises in year

In this model, year is used as fixed effect and cruise/jar as random factors, therefore testing if variability among years is larger than within-jar/cruise.

downsides: Some years have more cruises/samples than others, which can influence the precision Years are not considered directional, and this model will not show overall trends. Considering that the main goal is to identify changes in response to climate anomalies, this is not a problem.

## Uneven Sampling
Some diagnostic and informative tables for decision making, set up with eval = false to not run every time

checking number of cruises per months:

```{r eval=FALSE}
cruise_per_month <- data_date %>%
    group_by(month = month(base_date)) %>%
  summarise(n_cruises = n_distinct(base_date))
```


checking number of jars per cruise:

```{r eval=FALSE}
tot_jars_per_cruise <- data_date %>%
  group_by(base_date) %>%
  summarise(n_jars = n_distinct(jar))
```


checking number of jars per line per cruise

```{r eval=FALSE}
jars_per_line_per_cruise_not3 <- data_date %>%
  group_by(base_date, line) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop")
```


checking cruises with lines that have more or less than 3 samples:

```{r eval=FALSE}
jars_per_line_not3 <- data_date %>%
  group_by(base_date, line) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop") %>%
  filter(n_jars != 3)
```

Checking number of lines per cruise, and identifying missing lines

```{r eval=FALSE}
missing_lines <- data_date %>%
  group_by(base_date) %>%
  summarise(
    n_lines = n_distinct(line),
    missing_lines = paste(setdiff(c("2", "4", "6"), unique(line)), collapse = ", "),
    .groups = "drop")
```


checking stations of each line across the years, revealing multiple station per line, some N/A, some have ".", some "TT". conclusion: do not use station column as random factors

```{r eval=FALSE}
station_per_line <- df %>%
  group_by(line, station) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop") 
```


checking my counts:

```{r eval=FALSE}
jar_counts <- df %>%
  group_by(species_code, jar) %>%
  summarise(total_individuals = sum(tot_num, na.rm = TRUE), .groups = "drop") %>%
  arrange(species_code, jar)
```



# Size Calculation


Adding columns for width and length midpoint for each size range in each type:

```{r}
df <- df %>%
  mutate(length_mid = (length_l + length_h) / 2) %>%
  mutate(width_mid = (width_l + width_h) / 2) %>%
relocate(length_mid, .after = length_h) %>%
  relocate(width_mid, .after = width_h)
```

Calculating volume per each row by considering krill cone-shaped, using width_mid as diameter and length_mid as height:

```{r}
df <- df %>%
  mutate(volume = (1/3) * pi * (width_mid / 2)^2 * length_mid) %>%
  relocate(volume, .after = width_mid) %>%
  filter(!is.na(volume))
```

Calculating mean volume per jar: (weighted mean of volume instead of means of l and w first then calculating mean volume with mean l and w)
```{r}
df_vol <- df %>%
  group_by(species_code, jar, type, base_date, moci, line) %>%
  filter(!is.na(tot_num)) %>%
  summarise(volume_weighted = sum(volume * tot_num) / sum(tot_num),
    .groups = "drop")
```


# Figures

Creating a plot function for all species' adults:

```{r}
plot_adult <- function(data, species_name = "Adult") {

  data <- data %>%
    mutate(year = year(base_date))
  
  summary_vol <- data %>%
    group_by(year, type) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot() +

    # Volume (viridis, consistent)
    geom_line(data = summary_vol,
              aes(x = factor(year), y = volume_weighted,
                  color = type, group = type),
              linewidth = 1) +

    geom_point(data = summary_vol,
               aes(x = factor(year), y = volume_weighted,
                   color = type),
               size = 3) +

    # MOCI (neutral + dashed ALWAYS)
    geom_line(data = summary_moci,
              aes(x = factor(year),
                  y = (moci * scale_factor) + vol_mid,
                  group = 1),
              color = "grey30",
              linewidth = 1.2,
              linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +

    scale_color_viridis_d(option = "D", end = 0.8) +

labs(
  x = "Year",
  color = "Type",
  title = bquote(italic(.(species_name)))
) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```


Creating a plot function for all species' juveniles:
```{r}

plot_juv <- function(data, species_name = "juvenile") {

  data <- data %>%
    mutate(year = year(base_date))

  summary_vol <- data %>%
    group_by(year) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot(data, aes(x = factor(year))) +

    stat_summary(aes(y = volume_weighted, group = 1),
                 fun = mean, geom = "line",
                 linewidth = 1,
                 color = viridis::viridis(1)) +

    stat_summary(aes(y = volume_weighted, group = 1),
                 fun = mean, geom = "point",
                 size = 3,
                 color = viridis::viridis(1)) +

    # MOCI (scaled)
    stat_summary(aes(y = (moci * scale_factor) + vol_mid, group = 1),
                 fun = mean, geom = "line",
                 color = "grey30", linewidth = 1.2, linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +


    
    labs(
  x = "Year",
  title = bquote(italic(.(species_name)))
) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

## Euphausia pacifica

filtering species:

```{r}
epac_df <- df_vol %>%
  filter(species_code == "EPAC")
```

Creating separate dataframe for adults and juveniles:

```{r}
epac_adult <- epac_df %>%
  filter(type %in% c("f", "m"))

epac_juv <- epac_df %>%
  filter(type == "juv")
```

Creating plot for EPAC adult:
```{r}
plot_adult(epac_adult)
```


Creating plot for EPAC juveniles:
```{r}
plot_juv(epac_juv)
```



## Thysanoessa spinifera

filtering species:

```{r}
tspin_df <- df_vol %>%
  filter(species_code == "TSPIN")
```

Creating separate dataframe for adults and juveniles:

```{r}
tspin_adult <- tspin_df %>%
  filter(type %in% c("f", "m"))

tspin_juv <- tspin_df %>%
  filter(type == "juv")
```


```{r}
plot_adult(tspin_adult)
```


Creating plot for EPAC juveniles:
```{r}
plot_juv(tspin_juv)
```

## Nematocellis difficilis

filtering species:

```{r}
ndif_df <- df_vol %>%
  filter(species_code == "NDIF")
```

Creating separate dataframe for adults and juveniles:

```{r}
ndif_adult <- ndif_df %>%
  filter(type %in% c("f", "m"))

ndif_juv <- ndif_df %>%
  filter(type == "juv")
```


```{r}
plot_adult(ndif_adult)
```



Creating plot for EPAC juveniles:
```{r}
plot_juv(ndif_juv)
```



# comparing species

creating the df


plotting m

```{r}
plot_adultm_species_compare <- function(data, sex_filter = "m", title_label = "Adults") {

  data <- data %>%
    mutate(year = year(base_date)) %>%
    filter(type == sex_filter)

  # Volume summary (per species per year)
  summary_vol <- data %>%
    group_by(year, species_code) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  # MOCI summary (shared across species)
  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  # Scaling (same axis)
  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot() +

    # Species lines (viridis)
    geom_line(data = summary_vol,
              aes(x = factor(year),
                  y = volume_weighted,
                  color = species_code,
                  group = species_code),
              linewidth = 1) +

    geom_point(data = summary_vol,
               aes(x = factor(year),
                   y = volume_weighted,
                   color = species_code),
               size = 3) +

    # MOCI (shared, dashed)
    geom_line(data = summary_moci,
              aes(x = factor(year),
                  y = (moci * scale_factor) + vol_mid,
                  group = 1),
              color = "grey30",
              linewidth = 1.2,
              linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +

    scale_color_viridis_d(option = "D", end = 0.85) +

    labs(
      x = "Year",
      color = "Species",
      title = paste(title_label, "-", sex_filter)) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1))
}
```


```{r}
plot_adultm_species_compare(df_vol)
```


plotting f

```{r}
plot_adultf_species_compare <- function(data, sex_filter = "f", title_label = "Adults") {

  data <- data %>%
    mutate(year = year(base_date)) %>%
    filter(type == sex_filter)

  # Volume summary (per species per year)
  summary_vol <- data %>%
    group_by(year, species_code) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  # MOCI summary (shared across species)
  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  # Scaling (same axis)
  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot() +

    # Species lines (viridis)
    geom_line(data = summary_vol,
              aes(x = factor(year),
                  y = volume_weighted,
                  color = species_code,
                  group = species_code),
              linewidth = 1) +

    geom_point(data = summary_vol,
               aes(x = factor(year),
                   y = volume_weighted,
                   color = species_code),
               size = 3) +

    # MOCI (shared, dashed)
    geom_line(data = summary_moci,
              aes(x = factor(year),
                  y = (moci * scale_factor) + vol_mid,
                  group = 1),
              color = "grey30",
              linewidth = 1.2,
              linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +

    scale_color_viridis_d(option = "D", end = 0.85) +

    labs(
      x = "Year",
      color = "Species",
      title = paste(title_label, "-", sex_filter)) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1))
}
```



```{r}
plot_adultf_species_compare(df_vol)
```


# Maps

```{r}
df_map <- df %>%
  left_join(
    loc_data %>%
      select(ID_Jar, Lon_IN, Lat_IN),
    by = c("jar" = "ID_Jar"))
```



making a clean df:
```{r}
df_map_clean <- df_map %>%
  select(jar, cruise, Lat_IN, Lon_IN, base_date, line, station) %>%
  distinct() %>%
  filter(!is.na(Lat_IN), !is.na(Lon_IN))
```


creating a map: ADD BATHYMETRY LAYER, ADD FIXED TRANSECT LINES
CHECK:
line 2: 2141, 2142, 
line 4: 2131, 2134, 2137, 2140

```{r}

pal <- colorFactor(viridis(3, option = "D"), df_map_clean$line)


leaflet(df_map_clean) %>%
  addTiles() %>%  # base map
  addCircleMarkers(
    lng = ~Lon_IN,
    lat = ~Lat_IN,
    radius = 5,
    stroke = FALSE,
    fillOpacity = 0.8,
    color = ~pal(line),  # color by line
    label = ~paste("Jar:", jar, "<br>Cruise:", cruise, "<br>Line:", line)) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~line,  # must match palette
    title = "Line",
    opacity = 1
  )
```

# model with functions:

creating df function
```{r}
prep_species <- function(df_vol, species_name) {
  df_vol %>%
    filter(species_code == species_name) %>%
    mutate(
      year = lubridate::year(base_date),
      type = factor(type),
      line = factor(line),
      base_date = as.Date(base_date)
    ) %>%
    ungroup()
}


```

model function

```{r}
run_models <- function(df_vol) {

  # Year model
  model_year <- lmer(
    volume_weighted ~ factor(year) + type +
      (1 | base_date) + (1 | line),
    data = df_vol
  )
  
  # MOCI model
  model_moci <- lmer(
    volume_weighted ~ moci + type +
      (1 | base_date) + (1 | line),
    data = df_vol
  )
  
  # MOCI + type interaction
  model_moci_type <- lmer(
    volume_weighted ~ moci * type +
      (1 | base_date) + (1 | line),
    data = df_vol
  )
  
  return(list(
    year = model_year,
    moci = model_moci,
    moci_type = model_moci_type
  ))
}
```


summerizing models:

```{r}
summarize_models <- function(models, species_name) {
  
  cat("\n==============================\n")
  cat("Species:", species_name, "\n")
  cat("==============================\n")
  
  cat("\n--- YEAR EFFECT ---\n")
  print(anova(models$year))
  
  cat("\n--- MOCI EFFECT ---\n")
  print(summary(models$moci)$coefficients)
  
  cat("\n--- MOCI * TYPE ---\n")
  print(summary(models$moci_type)$coefficients)
}
```


running it for all species:

```{r}
epac_models <- prep_species(df_vol, "EPAC") %>% run_models()
summarize_models(epac_models, "EPAC")

tspin_models <- prep_species(df_vol, "TSPIN") %>% run_models()
summarize_models(tspin_models, "TSPIN")

ndif_models <- prep_species(df_vol, "NDIF") %>% run_models()
summarize_models(ndif_models, "NDIF")

```




# model

Running a linear mixed effects model for each species:

```{r}
epac_lme <- df_vol %>% filter(species_code == "EPAC") %>%
  mutate(year = year(base_date))

model_year <- lmer(volume_weighted ~ factor(year) + type + (1 | base_date) + (1 | line), data = epac_lme)

anova(model_year)
```
now testing moci:

```{r}
model_moci <- lmer(volume_weighted ~ moci + type +
                   (1 | base_date) + (1 | line),
                   data = epac_lme)

summary(model_moci)
```

moci inndividually in each type:

```{r}
model_moci_type <- lmer(volume_weighted ~ moci * type +
                        (1 | base_date) + (1 | line),
                        data = epac_lme)

summary(model_moci_type)
```
quick cheeky plot:

```{r}
ggplot(epac_lme, aes(x = moci, y = volume_weighted, color = type)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_classic()
```

