---
title: "Krill Size"
subtitle: "Changes in krill size in CCE"
author: Lala Movsesian
format: html
---

# Setup

loading packages and data

```{r}
library(here)
library(readxl)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest) 

raw_data <- read_excel(here::here("data", "raw", "krill_size.xlsx"))
moci_data <- read.csv(here::here("data", "raw", "CaliforniaMOCI_long.csv"))
```

### Date

Extracting year and month information from the cruise column and converting it to a convenient format. Creating base_date in the yyyy-mm-01 format to make sure it is treated as date in R. Used "cruise" column as reference rather than "date", since some sampling is done early in the next month (e.g. sampling date is 7/1 but it's from june cruise)

```{r}
data_date <- raw_data %>%

  mutate(base_date = make_date(
      year  = 2000 + as.numeric(substr(cruise, 4, 5)),
      month = as.numeric(substr(cruise, 6, 7)),
      day   = 1))%>% 
  relocate(base_date, .after = date)

```

excluding cruises from feb (n=1), apr (n=2), and oct (n=3)

```{r}
df <- data_date %>%
  filter(!month(base_date) %in% c(2, 4, 10))

```

### Life Stage

Changing all imm m to m, and imm f to f for consistency, since some samples include rows labeled as m in lower than 11mm ranges.

```{r}
df <- df %>%
  mutate(
    type = case_when(
      type == "imm m" ~ "m",
      type == "imm f" ~ "f",
      TRUE ~ type))
```

### Species

Filtering data: only including species of interest, only including juv, f and m

```{r}
df <- df %>%
  filter(species_name %in% c(
    "EUP Euphausia pacifica",
    "EUP Nematoscelis difficilis",
    "EUP Thysanoessa spinifera")) %>%
  
  # Keep only relevant types
  filter(type %in% c("juv", "f", "m"))
```

Adding code names for convenience

```{r}
df <- df %>%
  mutate(
    species_code = case_when(
      species_name == "EUP Euphausia pacifica" ~ "EPAC",
      species_name == "EUP Thysanoessa spinifera" ~ "TSPIN",
      species_name == "EUP Nematoscelis difficilis" ~ "NDIF")) %>%
  #move column
relocate(species_code, .after = species_name)
```

## MOCI

Joining seasonal MOCI from Northern California with df

```{r}
df <- df %>%
  mutate(
    Year = year(base_date),
    Month = month(base_date),
    Season = case_when(
      Month %in% c(4, 5, 6) ~ "AMJ",
      Month %in% c(7, 8, 9) ~ "JAS"))
    
df <- df %>%
  left_join(
    moci_data %>%
      select(Year, Season, North.California..38.42N.),
    by = c("Year", "Season")) %>%
  rename(moci = North.California..38.42N.)
```





# Analyses

Testing interannual variability using Linear Mixed Models because:

-   nested data (no need to collapse data to means)

-   not independent data (avoiding pseudoreplication and inflated significance)

-   considering uneven sampling effort

-   more suitable for different months and different numbers of cruises in year

In this model, year is used as fixed effect and cruise/jar as random factors, therefore testing if variability among years is larger than within-jar/cruise.

downsides: Some years have more cruises/samples than others, which can influence the precision Years are not considered directional, and this model will not show overall trends. Considering that the main goal is to identify changes in response to climate anomalies, this is not a problem.

## Uneven Sampling
Some diagnostic and informative tables for decision making, set up with eval = false to not run every time

checking number of cruises per months:

```{r eval=FALSE}
cruise_per_month <- data_date %>%
    group_by(month = month(base_date)) %>%
  summarise(n_cruises = n_distinct(base_date))
```


checking number of jars per cruise:

```{r eval=FALSE}
tot_jars_per_cruise <- data_date %>%
  group_by(base_date) %>%
  summarise(n_jars = n_distinct(jar))
```


checking number of jars per line per cruise

```{r eval=FALSE}
jars_per_line_per_cruise_not3 <- data_date %>%
  group_by(base_date, line) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop")
```


checking cruises with lines that have more or less than 3 samples:

```{r eval=FALSE}
jars_per_line_not3 <- data_date %>%
  group_by(base_date, line) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop") %>%
  filter(n_jars != 3)
```

Checking number of lines per cruise, and identifying missing lines

```{r eval=FALSE}
missing_lines <- data_date %>%
  group_by(base_date) %>%
  summarise(
    n_lines = n_distinct(line),
    missing_lines = paste(setdiff(c("2", "4", "6"), unique(line)), collapse = ", "),
    .groups = "drop")
```


checking stations of each line across the years, revealing multiple station per line, some N/A, some have ".", some "TT". conclusion: do not use station column as random factors

```{r eval=FALSE}
station_per_line <- df %>%
  group_by(line, station) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop") 
```


# Size Calculation

Calculating volume of krill considering each to be the shape of a cone and using cone volume formula. for each size range, d is median weight and h is median length. 

Adding columns for median weigth (d) and median length (h)
Using mid point size per row with length data only per type (juv/f/m) per jar per size range

```{r}
df_lmm <- df %>%
  mutate(length_mid = (length_l + length_h) / 2) %>%
relocate(length_mid, .after = length_h)
```

Calculating one size per type per jar

```{r}
df_lmm <- df_lmm %>%
  group_by(species_code, jar, type, base_date, moci) %>%
  summarise(length_weighted = sum(length_mid * tot_num) / sum(tot_num),
    .groups = "drop")
```

## Euphausia pacifica

filtering species and N/As:

```{r}
epac_df <- df_lmm %>%
  filter(species_code == "EPAC") %>%
  filter(!is.na(length_weighted))
```

Creating separate dataframe for adults and juveniles:

```{r}
epac_adult <- epac_df %>%
  filter(type %in% c("f", "m"))

epac_juv <- epac_df %>%
  filter(type == "juv")
```

### Figures

figure showing the size of adults over the years:

TEST
```{r}
epac_adult <- epac_adult %>%
  mutate(year = year(base_date))

# ranges
len_min <- min(epac_adult$length_weighted, na.rm = TRUE)
len_max <- max(epac_adult$length_weighted, na.rm = TRUE)
len_mid <- mean(c(len_min, len_max))

moci_max <- max(abs(epac_adult$moci), na.rm = TRUE)

# scaling factor (fit MOCI into length range nicely)
scale_factor <- (len_max - len_min) / (2 * moci_max)

ggplot(epac_adult, aes(x = factor(year))) +

  # jitter
  #geom_jitter(aes(y = length_weighted, color = type), width = 0.2, alpha = 0.4, size = 1) +

  # means
  stat_summary(aes(y = length_weighted, color = type, group = type),
               fun = mean, geom = "line", linewidth = 1) +

  stat_summary(aes(y = length_weighted, color = type),
               fun = mean, geom = "point", size = 3) +

  # MOCI (centered + scaled + shifted)
  stat_summary(aes(y = (moci * scale_factor) + len_mid, group = 1),
               fun = mean, geom = "line",
               color = "black", linewidth = 1.2, linetype = "dashed") +

  scale_y_continuous(
    name = "Weighted mean body length (mm)",
    sec.axis = sec_axis(~ (. - len_mid) / scale_factor,
                        name = "MOCI")
  ) +

  scale_color_manual(values = c("f" = "#E64B35", "m" = "#4DBBD5")) +

  labs(x = "Year", color = "Type") +

  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1))
```



figure showing the size of juveniles over the years:

```{r}
ggplot(epac_juv, aes(x = factor(year(base_date)), y = length_weighted)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  labs(x = "Year", y = "Weighted mean body length (mm)") +

  theme_minimal(base_size = 13) +
theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

### Model

Running model with epac adults: first checking variables

```{r}
epac_adult <- epac_adult %>%
  mutate(year = factor(year(base_date)),
    cruise = factor(base_date),
    jar = factor(jar),
    type = factor(type))
```

running mixed models:

```{r}
model_epac <- lmer(length_weighted ~ year + type + (1 | cruise/jar),
  data = epac_adult)

summary(model_epac)
```

running anova

```{r}
anova(model_epac)
```

making a plot for years:

```{r}
library(emmeans)
year_means <- emmeans(model_epac, ~ year)
plot(year_means)
```

## Thysanoessa spinifera

filtering species and N/As:

```{r}
tspin_df <- df_lmm %>%
  filter(species_code == "TSPIN") %>%
  filter(!is.na(length_weighted))
```

Creating separate dataframe for adults and juveniles:

```{r}
tspin_adult <- tspin_df %>%
  filter(type %in% c("f", "m"))

tspin_juv <- tspin_df %>%
  filter(type == "juv")
```

### Figures

figure showing the size of adults over the years:

```{r}
ggplot(tspin_adult, aes(x = factor(year(base_date)), y = length_weighted, color = type)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  scale_color_manual(values = c("f" = "#E64B35", "m" = "#4DBBD5")) +

  labs(x = "Year", y = "Weighted mean body length (mm)", color = "Type") +

  theme_minimal(base_size = 13) +

  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

figure showing the size of juveniles over the years:

```{r}
ggplot(tspin_juv, aes(x = factor(year(base_date)), y = length_weighted)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  labs(x = "Year", y = "Weighted mean body length (mm)") +

  theme_minimal(base_size = 13) +
theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

## Nematoscelis difficilis

filtering species and N/As:

```{r}
ndif_df <- df_lmm %>%
  filter(species_code == "NDIF") %>%
  filter(!is.na(length_weighted))
```

Creating separate dataframe for adults and juveniles:

```{r}
ndif_adult <- ndif_df %>%
  filter(type %in% c("f", "m"))

ndif_juv <- ndif_df %>%
  filter(type == "juv")
```

### Figures

figure showing the size of adults over the years:

```{r}
ggplot(ndif_adult, aes(x = factor(year(base_date)), y = length_weighted, color = type)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  scale_color_manual(values = c("f" = "#E64B35", "m" = "#4DBBD5")) +

  labs(x = "Year", y = "Weighted mean body length (mm)", color = "Type") +

  theme_minimal(base_size = 13) +

  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

figure showing the size of juveniles over the years:

```{r}
ggplot(ndif_juv, aes(x = factor(year(base_date)), y = length_weighted)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  labs(x = "Year", y = "Weighted mean body length (mm)") +

  theme_minimal(base_size = 13) +
theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```
