---
title: "Krill Size"
subtitle: "Changes in krill size in CCE"
author: Lala Movsesian
format: html
toc: true
toc-depth: 3
---

# Setup

loading packages and data

```{r}
library(here)
library(readxl)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest) 
library(viridis)
library(leaflet)
library(broom.mixed)
library(DT)
library(mgcv)
library(purrr)
library(patchwork)
library(ggeffects)
library(gratia)



raw_data <- read_excel(here("data", "raw", "krill_size.xlsx"))
moci_data <- read.csv(here("data", "raw", "CaliforniaMOCI_long.csv"))
loc_data <- read_excel(here("data","raw", "locdata.xlsx"))
climate_raw <- read_excel(here("data","raw", "climate_ocean_indices.xlsx"))

```

# Creating main dataframe

## Filtering data

Extracting year and month information from the cruise column and converting it to a convenient format. Creating base_date in the yyyy-mm-01 format to make sure it is treated as date in R. Used "cruise" column as reference rather than "date", since some sampling is done early in the next month (e.g. sampling date is 7/1 but it's from june cruise)

```{r}
data_all <- raw_data %>%
  mutate(base_date = make_date(
      year  = 2000 + as.numeric(substr(cruise, 4, 5)),
      month = as.numeric(substr(cruise, 6, 7)),
      day   = 1))%>% 
  relocate(base_date, .after = date)

```

Excluding cruises from feb (n=1), apr (n=2), and oct (n=3).Changing all imm m to m, and imm f to f for consistency, since some samples include rows labeled as m in lower than 11mm ranges, and filtering for desired life stages (excluding zoea). Only including numerically dominant species (excluding euphausiidae), adding species code names for convenience:

```{r}
data_krill <- data_all %>%
  filter(!month(base_date) %in% c(2, 4, 10)) %>%  
  mutate(
    type = case_when(
      type == "imm m" ~ "m",
      type == "imm f" ~ "f",
      TRUE ~ type)) %>%
  filter(type %in% c("juv", "f", "m")) %>%
  filter(species_name %in% c(
    "EUP Euphausia pacifica",
    "EUP Nematoscelis difficilis",
    "EUP Thysanoessa spinifera")) %>%
  mutate(
    species_code = case_when(
      species_name == "EUP Euphausia pacifica" ~ "EPAC",
      species_name == "EUP Thysanoessa spinifera" ~ "TSPIN",
      species_name == "EUP Nematoscelis difficilis" ~ "NDIF")) %>%

relocate(species_code, .after = species_name)
```

## Adding MOCI

Joining seasonal MOCI from Northern California with df

```{r}
data_krill <- data_krill %>%
  mutate(
    Year = year(base_date),
    Month = month(base_date),
    Season = case_when(
      Month %in% c(4, 5, 6) ~ "AMJ",
      Month %in% c(7, 8, 9) ~ "JAS"))
    
data_moci <- data_krill %>%
  left_join(
    moci_data %>%
      select(Year, Season, North.California..38.42N.),
    by = c("Year", "Season")) %>%
  rename(moci = North.California..38.42N.) %>%
  rename(season = Season) %>%
  select(-Year, -Month) %>%
  relocate(season, .after = base_date)
```

## Size Calculation

Adding columns for width and length midpoint for each size range in each type:

```{r}
df <- data_moci %>%
  mutate(length_mid = (length_l + length_h) / 2) %>%
  mutate(width_mid = (width_l + width_h) / 2) %>%
relocate(length_mid, .after = length_h) %>%
  relocate(width_mid, .after = width_h)
```

Calculating volume per each row by considering krill cone-shaped, using width_mid as diameter and length_mid as height:

```{r}
df <- df %>%
  mutate(volume = (1/3) * pi * (width_mid / 2)^2 * length_mid) %>%
  relocate(volume, .after = width_mid) %>%
  filter(!is.na(volume))
```

Calculating mean volume per jar: (weighted mean of volume instead of means of l and w first then calculating mean volume with mean l and w)

```{r}
df_vol <- df %>%
  group_by(species_code, jar, type, base_date, moci, line, cruise) %>%
  filter(!is.na(tot_num)) %>%
  summarise(volume_weighted = sum(volume * tot_num) / sum(tot_num),
    .groups = "drop")
```

# Analysis - MOCI

## Linear Mixed Effects Model

creating function:

```{r}
prep_species <- function(df_vol, species_name) {
  df_vol %>%
    filter(species_code == species_name) %>%
    mutate(
      year = lubridate::year(base_date),
      type = factor(type),
      line = factor(line),
      base_date = as.Date(base_date)) %>%
    ungroup()}

run_models <- function(df_vol) {
  # Year model
  model_year <- lmer(
    volume_weighted ~ factor(year) + type +
      (1 | base_date) + (1 | line),
    data = df_vol)
  
  # MOCI model
  model_moci <- lmer(
    volume_weighted ~ moci + type +
      (1 | base_date) + (1 | line),
    data = df_vol)
  
  # MOCI + m/f/juv interaction
  model_moci_type <- lmer(
    volume_weighted ~ moci * type +
      (1 | base_date) + (1 | line),
    data = df_vol)
  
  return(list(
    year = model_year,
    moci = model_moci,
    moci_type = model_moci_type))}
```

running it for all species:

```{r}
epac_models <- prep_species(df_vol, "EPAC") %>% run_models()

tspin_models <- prep_species(df_vol, "TSPIN") %>% run_models()

ndif_models <- prep_species(df_vol, "NDIF") %>% run_models()
```

generating tables:

```{r}
#creating function 
extract_results <- function(models, species_name) {
  broom.mixed::tidy(models$moci_type, effects = "fixed") %>%
    mutate(species = species_name)}

#running function for each species 
epac_res  <- extract_results(epac_models, "EPAC")
tspin_res <- extract_results(tspin_models, "TSPIN")
ndif_res  <- extract_results(ndif_models, "NDIF")

#combining all 3 tables vertically
all_results <- bind_rows(epac_res, tspin_res, ndif_res)

#keeping desired columns only
clean_table <- all_results %>%
  select(species, term, estimate, std.error, statistic, p.value) %>%
  #rounding to 3 decimals, and keeping 3 significant digits for p value
  mutate(
    across(c(estimate, std.error, statistic), round, 3),
    p.value = signif(p.value, 3)) %>%
  mutate(term = recode(term,
    "(Intercept)" = "Intercept",
    "moci" = "MOCI",
    "typejuv" = "Juvenile",
    "typem" = "Male",
    "moci:typejuv" = "MOCI × Juvenile",
    "moci:typem" = "MOCI × Male"))

#creating interactive table
datatable(clean_table,
          options = list(pageLength = 18),
          rownames = FALSE)

```

Plot to show and compare slopes:

```{r}
ggplot(df_vol %>% filter(species_code %in% c("EPAC","NDIF","TSPIN")), 
       aes(x = moci, y = volume_weighted, color = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~species_code) +
  theme_classic()
```

## Linear trends

```{r}
run_lme_trend <- function(df_vol) {

  data_directional <- df_vol %>%
    mutate(
      year = lubridate::year(base_date),
      year_scaled = scale(year, center = TRUE, scale = FALSE))

  model <- lmer(
    volume_weighted ~ year_scaled + moci + type + (1 | year),
    data = data_directional)

}
```

running models:

```{r}
epac_trend <- prep_species(df_vol, "EPAC") %>% run_lme_trend()

tspin_trend <- prep_species(df_vol, "TSPIN") %>% run_lme_trend()

ndif_trend <- prep_species(df_vol, "NDIF") %>% run_lme_trend()
```

showing results: FIX THIS PART

```{r}
print(epac_trend)
print(tspin_trend)
print(ndif_trend)
```

## GAM

testing GAM:

```{r}
run_gam_trend <- function(df_vol) {

  data_gam <- df_vol %>%
    mutate(year = lubridate::year(base_date))

  model <- gam(
    volume_weighted ~ s(year, k = 10) + moci + type,
    data = data_gam,
    method = "REML")

  return(model)}
```

```{r}
model_epac_gam <- run_gam_trend(prep_species(df_vol, "EPAC"))
summary(model_epac_gam)
plot(model_epac_gam, shade = TRUE)

model_tspin_gam <- run_gam_trend(prep_species(df_vol, "TSPIN"))
summary(model_tspin_gam)
plot(model_tspin_gam, shade = TRUE)

model_ndif_gam <- run_gam_trend(prep_species(df_vol, "NDIF"))
summary(model_ndif_gam)
plot(model_ndif_gam, shade = TRUE)
```

### Poster figures

#### 1. gerneral size bar chart:

```{r}
life_stage_summary <- df_vol %>%
  group_by(species_code, type) %>%
  summarise(
    mean_size = mean(volume_weighted, na.rm = TRUE),
    se = sd(volume_weighted, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

life_stage_summary <- life_stage_summary %>%
  mutate(
    species = recode(species_code,
                     "EPAC" = "E. pacifica",
                     "TSPIN" = "T. spinifera",
                     "NDIF" = "N. difficilis"),
    type = recode(type,
                  "f" = "Female",
                  "m" = "Male",
                  "juv" = "Juvenile")
  )

life_stage_summary$type <- factor(
  life_stage_summary$type,
  levels = c("Juvenile", "Male", "Female")
)


ggplot(life_stage_summary,
       aes(x = species, y = mean_size, fill = type)) +
  
  geom_col(position = position_dodge(width = 0.7),
           width = 0.6) +
  
  geom_errorbar(aes(ymin = mean_size - se,
                    ymax = mean_size + se),
                position = position_dodge(width = 0.7),
                width = 0.2) +
  
  scale_fill_viridis_d(option = "D", end = 0.8) +
  
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 5)  # ~2–3 gridlines instead of 10
  ) +
  
  labs(
    x = "Species",
    y = "Mean body volume (mm³)",
    fill = "Life stage",
    title = "Body size differences across species and life stages"
  ) +
  
  theme_minimal(base_size = 14) +
  
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),  # remove tiny gridlines
    panel.grid.major = element_line(linewidth = 0.4),
    axis.text.x = element_text(face = "italic")
  )
```


#### 2. species relative abundance pie chart
```{r}

species_sum <- df %>%
  group_by(species_code) %>%
  summarise(total_n = sum(tot_num, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    perc = total_n / sum(total_n) * 100,
    label = paste0(round(perc, 1), "%")
  )

species_sum$species_code <- factor(
  species_sum$species_code,
  levels = c("EPAC", "TSPIN", "NDIF")
)

ggplot(species_sum, aes(x = "", y = total_n, fill = species_code)) +
  
  geom_col(width = 1, color = "white") +
  
  coord_polar(theta = "y") +
  
  scale_fill_viridis_d(option = "D", end = 0.85) +
  
  
  labs(
    fill = "Species",
    title = "Species composition of krill samples"
  ) +
  
  theme_void() +
  
  theme(plot.title = element_text(face = "bold"))
```






#### 4. standardized effects plot (with gam)

```{r}
# EPAC
epac_eff <- ggpredict(model_epac_gam, terms = "moci")

# NDIF
ndif_eff <- ggpredict(model_ndif_gam, terms = "moci")

# tspin
tspin_eff <- ggpredict(model_tspin_gam, terms = "moci")

epac_eff$species <- "E. pacifica"
ndif_eff$species <- "N. difficilis"
tspin_eff$species <- "T. spinifera"

eff_all <- bind_rows(epac_eff, ndif_eff, tspin_eff)


eff_all2 <- eff_all %>%
  group_by(species) %>%
  mutate(
    predicted_centered = predicted - predicted[which.min(abs(x - mean(x)))],
    conf.low_centered = conf.low - predicted[which.min(abs(x - mean(x)))],
    conf.high_centered = conf.high - predicted[which.min(abs(x - mean(x)))]
  ) %>%
  ungroup()


ggplot(eff_all2,
       aes(y = predicted_centered, x = x, color = species)) +
  
  # main lines
  geom_line(linewidth = 1.4) +
  
  # subtle endpoints (start + end of each curve)
  geom_point(
    data = eff_all2 %>%
      group_by(species) %>%
      slice(c(1, n())) %>%
      ungroup(),
    aes(y = predicted_centered, x = x, color = species),
    size = 2,
    alpha = 0.6
  ) +
  
  # reference line (no climate effect baseline)
  geom_hline(yintercept = mean(eff_all2$x), 
             linetype = "dashed", alpha = 0.4) +

  
  scale_color_viridis_d(option = "D", end = 0.85) +
  
  labs(
    y = "Deviation in predicted body size",
    x = "MOCI (climate conditions)",
    title = "Species-specific responses of krill body size to climate variability"
  ) +
  
  theme_minimal(base_size = 14) +
  
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```






#### 5. linear vs ninlinear

```{r}


# Extract smooths for each species


epac_smooth <- smooth_estimates(model_epac_gam, smooth = "s(year)") %>%
  mutate(species = "E. pacifica")

ndif_smooth <- smooth_estimates(model_ndif_gam, smooth = "s(year)") %>%
  mutate(species = "N. difficilis")

tspin_smooth <- smooth_estimates(model_tspin_gam, smooth = "s(year)") %>%
  mutate(species = "T. spinifera")


# Combine + add CI


smooth_all <- bind_rows(epac_smooth, ndif_smooth, tspin_smooth) %>%
  mutate(
    .lower = .estimate - 1.96 * .se,
    .upper = .estimate + 1.96 * .se
  )

# order species (important for visual story)
smooth_all$species <- factor(
  smooth_all$species,
  levels = c("N. difficilis", "E. pacifica", "T. spinifera")
)


# Plot


ggplot(smooth_all, aes(x = year, y = .estimate)) +
  
  # confidence ribbon
  geom_ribbon(aes(ymin = .lower, ymax = .upper),
              fill = "grey70",
              alpha = 0.4) +
  
  # smooth line
  geom_line(linewidth = 1.2, color = "black") +
  
  # zero reference
  geom_hline(yintercept = 0,
             linetype = "dashed",
             alpha = 0.5) +
  
  # facet by species
  facet_wrap(~ species, ncol = 1, scales = "fixed") +
  
  labs(
    x = "Year",
    y = "Effect on body size (relative)",
    title = "Temporal patterns in krill body size"
  ) +
  
  theme_minimal(base_size = 14) +
  
  theme(
    strip.text = element_text(face = "italic", size = 13),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```

#### 6. main effects
```{r}
effect_df <- bind_rows(
  
  tidy(ndif_trend, effects = "fixed") %>%
    filter(term == "moci") %>%
    mutate(species = "N. difficilis"),
  
  tidy(epac_trend, effects = "fixed") %>%
    filter(term == "moci") %>%
    mutate(species = "E. pacifica"),
  
  tidy(tspin_trend, effects = "fixed") %>%
    filter(term == "moci") %>%
    mutate(species = "T. spinifera")
) %>%
  select(species, estimate, std.error) %>%
  rename(se = std.error)


effect_df$species <- factor(effect_df$species,
                            levels = c("N. difficilis",
                                       "E. pacifica",
                                       "T. spinifera"))

ggplot(effect_df, aes(x = species, y = estimate, fill = species)) +
  
  geom_col(width = 0.6, alpha = 0.9) +
  
  geom_errorbar(
    aes(ymin = estimate - se, ymax = estimate + se),
    width = 0.15,
    linewidth = 0.9
  ) +
  
  # zero reference line = "no effect"
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.8, alpha = 0.6) +
  
  scale_fill_viridis_d(option = "D", end = 0.85) + 
  
  labs(
    x = "",
    y = "Effect of MOCI on body size",
    title = "Species-specific sensitivity to climate variability"
  ) +
  
  theme_minimal(base_size = 14) +
  
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

#### anomalies

```{r}
prep_anomaly_data <- function(df, species_name) {
  
  df %>%
    filter(species_code == species_name) %>%
    mutate(year = year(base_date)) %>%
    
    group_by(year) %>%
    summarise(
      size_mean = mean(volume_weighted, na.rm = TRUE),
      moci_mean = mean(moci, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    
    mutate(
      size_anomaly = size_mean - mean(size_mean, na.rm = TRUE),
      moci_anomaly = moci_mean - mean(moci_mean, na.rm = TRUE)
    )
}
```

```{r}
epac_anom <- prep_anomaly_data(df_vol, "EPAC")
```

plotting bars:

```{r}
ggplot(epac_anom, aes(x = factor(year))) +
  
  # size anomaly (flip sign so it mirrors MOCI)
  geom_col(aes(y = -size_anomaly), fill = "steelblue", alpha = 0.8) +
  
  # MOCI anomaly
  geom_col(aes(y = moci_anomaly), fill = "firebrick", alpha = 0.6) +
  
  geom_hline(yintercept = 0, color = "black") +
  
  labs(
    x = "Year",
    y = "Anomaly (centered)",
    title = expression(italic("E. pacifica") ~ "size vs climate anomalies"),
    caption = "Blue = size anomaly (inverted); Red = MOCI anomaly"
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

another look:

```{r}
ggplot(epac_anom, aes(x = year)) +
  
  geom_col(aes(y = size_anomaly), fill = "steelblue", alpha = 0.7) +
  
  geom_line(aes(y = moci_anomaly), linewidth = 1.2) +
  geom_point(aes(y = moci_anomaly), size = 2) +
  
  geom_hline(yintercept = 0) +
  
  theme_minimal()
```

```{r}
ndif_anom <- prep_anomaly_data(df_vol, "NDIF")
```

plotting bars:

```{r}
ggplot(ndif_anom, aes(x = factor(year))) +
  
  # size anomaly (flip sign so it mirrors MOCI)
  geom_col(aes(y = -size_anomaly), fill = "steelblue", alpha = 0.8) +
  
  # MOCI anomaly
  geom_col(aes(y = moci_anomaly), fill = "firebrick", alpha = 0.6) +
  
  geom_hline(yintercept = 0, color = "black") +
  
  labs(
    x = "Year",
    y = "Anomaly (centered)",
    title = expression(italic("N. difficilis") ~ "size vs climate anomalies"),
    caption = "Blue = size anomaly (inverted); Red = MOCI anomaly"
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```



```{r}
tspin_anom <- prep_anomaly_data(df_vol, "TSPIN")
```

plotting bars:

```{r}
ggplot(tspin_anom, aes(x = factor(year))) +
  
  # size anomaly (flip sign so it mirrors MOCI)
  geom_col(aes(y = -size_anomaly), fill = "steelblue", alpha = 0.8) +
  
  # MOCI anomaly
  geom_col(aes(y = moci_anomaly), fill = "firebrick", alpha = 0.6) +
  
  geom_hline(yintercept = 0, color = "black") +
  
  labs(
    x = "Year",
    y = "Anomaly (centered)",
    title = expression(italic("T. spinifera") ~ "size vs climate anomalies"),
    caption = "Blue = size anomaly (inverted); Red = MOCI anomaly"
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```






# Analysis - other indices

organizing the data: cleaning up the names, changing all cruise names to ACC format

```{r}
climate <- climate_raw %>%
  filter(!is.na(cruise)) %>%
  rename_with(~ str_replace(., "-(\\d+)", "_lag\\1")) %>%
    mutate(cruise = case_when(
    cruise == "ACM1705" ~ "ACC1705", 
    cruise == "ACD1807" ~ "ACC1807",
    cruise == "ACM2205" ~ "ACC2205", 
    cruise == "ACD2206" ~ "ACC2206",
    
    TRUE ~ cruise))
```

adding lags in a separate column

```{r}
df_climate <- climate %>% 
  select(cruise, SOI, SOI_lag1, SOI_lag2, SOI_lag3, PDO, PDO_lag1, PDO_lag2, PDO_lag3, NPGO, NPGO_lag1, NPGO_lag2, NPGO_lag3, CUTI, CUTI_lag1, CUTI_lag2, CUTI_lag3, BEUTI, BEUTI_lag1, BEUTI_lag2, BEUTI_lag3, UI, UI_lag1, UI_lag2, UI_lag3, SEFI_SST_mean, SEFI_SST_anomalies, SEFI_SST_mean_lag1, SEFI_SST_anomalies_lag1, SEFI_SST_mean_lag2, SEFI_SST_anomalies_lag2, SEFI_SST_mean_lag3, SEFI_SST_anomalies_lag3)  %>%
  pivot_longer(
    cols = -cruise,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    index = str_extract(variable, "^[^_]+"),     # PDO, SOI, etc.
    lag = str_extract(variable, "(?<=lag)\\d+")  # 1, 2, 3
  ) %>%
  mutate(
    lag = ifelse(is.na(lag), 0, as.numeric(lag))
  ) %>%

  mutate(index = case_when(
    index == "SEFI" ~ "SST", TRUE ~ index))

```

model prep:

```{r}
df_indices <- df_vol %>%
  left_join(df_climate, by = "cruise")
```

model core function:

```{r}
run_model_species <- function(data, species_name, index_name, lag_value) {
  
  df_sub <- data %>%
    filter(
      species_code == species_name,
      index == index_name,
      lag == lag_value
    ) %>%
    filter(!is.na(value), !is.na(volume_weighted))
  
  # guards
  if (nrow(df_sub) < 5) return(NULL)
  if (n_distinct(df_sub$type) < 2) return(NULL)
  
  model <- tryCatch(
    lmer(
      volume_weighted ~ value * type +
        (1 | jar),
      data = df_sub
    ),
    error = function(e) return(NULL)
  )
  
  return(model)
}
```

running models

```{r}
species_list <- unique(df_indices$species_code)
indices <- unique(df_indices$index)
lags <- unique(df_indices$lag)
```

```{r}
species_results <- expand_grid(
  species = species_list,
  index = indices,
  lag = lags
) %>%
  mutate(
    model = pmap(
      list(species, index, lag),
      ~run_model_species(df_indices, ..1, ..2, ..3)
    )
  ) %>%
  filter(!map_lgl(model, is.null)) %>%
  mutate(
    tidy = map(model, ~broom.mixed::tidy(.x, effects = "fixed"))
  ) %>%
  select(-model) %>%
  unnest(tidy)
```

making a clean table for climate results:

```{r}
clean_species_table <- species_results %>%
  select(species, index, lag, term, estimate, std.error, statistic, p.value) %>%
  mutate(
    across(c(estimate, std.error, statistic), round, 3),
    p.value = signif(p.value, 3)
  ) %>%
  mutate(term = recode(term,
    "(Intercept)" = "Intercept",
    "value" = "Climate Index",
    "typejuv" = "Juvenile",
    "typem" = "Male",
    "value:typejuv" = "Climate × Juvenile",
    "value:typem" = "Climate × Male"
  ))
```

```{r}

datatable(
  clean_species_table,
  options = list(pageLength = 20),
  rownames = FALSE
)
```

```{r}
clean_species_table %>%
  filter(term == "Climate Index") %>%
  group_by(species, index) %>%
  summarise(mean_effect = mean(abs(estimate)), .groups = "drop") %>%
  arrange(desc(mean_effect))
```

three pannel figure:

```{r eval=FALSE}
top_indices <- importance_table %>%
  slice_head(n = 3) %>%
  pull(index)
```

pannel A:

```{r eval=FALSE}
p1 <- clim_main %>%
  filter(index %in% top_indices) %>%
  ggplot(aes(x = lag, y = estimate, color = index, group = index)) +
  geom_line(linewidth = 1) +
  geom_point() +
  theme_classic() +
  labs(title = "A. Climate effect across lags",
       y = "Effect size", x = "Lag")
```

```{r eval=FALSE}
p2 <- clim_main %>%
  filter(index %in% top_indices) %>%
  ggplot(aes(x = lag, y = index, fill = estimate)) +
  geom_tile() +
  scale_fill_gradient2() +
  theme_classic() +
  labs(title = "B. Index × lag response heatmap",
       x = "Lag", y = "Index")
```

```{r eval=FALSE}
p3 <- importance_table %>%
  filter(index %in% top_indices) %>%
  ggplot(aes(x = reorder(index, mean_effect), y = mean_effect)) +
  geom_col() +
  coord_flip() +
  theme_classic() +
  labs(title = "C. Climate driver importance",
       x = "Index", y = "Mean |effect|")
```

```{r eval=FALSE}
(p1 | p2) / p3
```

# Visualization

Creating a plot function for all species' adults:

```{r}
plot_adult <- function(data, species_name = "Adult") {

  data <- data %>%
    mutate(year = year(base_date))
  
  summary_vol <- data %>%
    group_by(year, type) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot() +

    # Volume (viridis, consistent)
    geom_line(data = summary_vol,
              aes(x = factor(year), y = volume_weighted,
                  color = type, group = type),
              linewidth = 1) +

    geom_point(data = summary_vol,
               aes(x = factor(year), y = volume_weighted,
                   color = type),
               size = 3) +

    # MOCI (neutral + dashed ALWAYS)
    geom_line(data = summary_moci,
              aes(x = factor(year),
                  y = (moci * scale_factor) + vol_mid,
                  group = 1),
              color = "grey30",
              linewidth = 1.2,
              linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +

    scale_color_viridis_d(option = "D", end = 0.8) +

labs(
  x = "Year",
  color = "Type",
  title = bquote(italic(.(species_name)))
) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

Creating a plot function for all species' juveniles:

```{r}

plot_juv <- function(data, species_name = "juvenile") {

  data <- data %>%
    mutate(year = year(base_date))

  summary_vol <- data %>%
    group_by(year) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot(data, aes(x = factor(year))) +

    stat_summary(aes(y = volume_weighted, group = 1),
                 fun = mean, geom = "line",
                 linewidth = 1,
                 color = viridis::viridis(1)) +

    stat_summary(aes(y = volume_weighted, group = 1),
                 fun = mean, geom = "point",
                 size = 3,
                 color = viridis::viridis(1)) +

    # MOCI (scaled)
    stat_summary(aes(y = (moci * scale_factor) + vol_mid, group = 1),
                 fun = mean, geom = "line",
                 color = "grey30", linewidth = 1.2, linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +


    
    labs(
  x = "Year",
  title = bquote(italic(.(species_name)))
) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

## Euphausia pacifica

filtering species:

```{r}
epac_df <- df_vol %>%
  filter(species_code == "EPAC")
```

Creating separate dataframe for adults and juveniles:

```{r}
epac_adult <- epac_df %>%
  filter(type %in% c("f", "m"))

epac_juv <- epac_df %>%
  filter(type == "juv")
```

Creating plot for EPAC adult:

```{r}
plot_adult(epac_adult)
```

Creating plot for EPAC juveniles:

```{r}
plot_juv(epac_juv)
```

## Thysanoessa spinifera

filtering species:

```{r}
tspin_df <- df_vol %>%
  filter(species_code == "TSPIN")
```

Creating separate dataframe for adults and juveniles:

```{r}
tspin_adult <- tspin_df %>%
  filter(type %in% c("f", "m"))

tspin_juv <- tspin_df %>%
  filter(type == "juv")
```

```{r}
plot_adult(tspin_adult)
```

Creating plot for EPAC juveniles:

```{r}
plot_juv(tspin_juv)
```

## Nematocellis difficilis

filtering species:

```{r}
ndif_df <- df_vol %>%
  filter(species_code == "NDIF")
```

Creating separate dataframe for adults and juveniles:

```{r}
ndif_adult <- ndif_df %>%
  filter(type %in% c("f", "m"))

ndif_juv <- ndif_df %>%
  filter(type == "juv")
```

```{r}
plot_adult(ndif_adult)
```

Creating plot for NDIF juveniles:

```{r}
plot_juv(ndif_juv)
```

## comparing species

plot for males:

```{r}
plot_adultm_species_compare <- function(data, sex_filter = "m", title_label = "Adults") {

  data <- data %>%
    mutate(year = year(base_date)) %>%
    filter(type == sex_filter)

  # Volume summary (per species per year)
  summary_vol <- data %>%
    group_by(year, species_code) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  # MOCI summary (shared across species)
  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  # Scaling (same axis)
  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot() +

    # Species lines (viridis)
    geom_line(data = summary_vol,
              aes(x = factor(year),
                  y = volume_weighted,
                  color = species_code,
                  group = species_code),
              linewidth = 1) +

    geom_point(data = summary_vol,
               aes(x = factor(year),
                   y = volume_weighted,
                   color = species_code),
               size = 3) +

    # MOCI (shared, dashed)
    geom_line(data = summary_moci,
              aes(x = factor(year),
                  y = (moci * scale_factor) + vol_mid,
                  group = 1),
              color = "grey30",
              linewidth = 1.2,
              linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +

    scale_color_viridis_d(option = "D", end = 0.85) +

    labs(
      x = "Year",
      color = "Species",
      title = paste(title_label, "-", sex_filter)) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r}
plot_adultm_species_compare(df_vol)
```

plot for females:

```{r}
plot_adultf_species_compare <- function(data, sex_filter = "f", title_label = "Adults") {

  data <- data %>%
    mutate(year = year(base_date)) %>%
    filter(type == sex_filter)

  # Volume summary (per species per year)
  summary_vol <- data %>%
    group_by(year, species_code) %>%
    summarise(volume_weighted = mean(volume_weighted, na.rm = TRUE),
              .groups = "drop")

  # MOCI summary (shared across species)
  summary_moci <- data %>%
    group_by(year) %>%
    summarise(moci = mean(moci, na.rm = TRUE),
              .groups = "drop")

  # Scaling (same axis)
  vol_min <- min(summary_vol$volume_weighted, na.rm = TRUE)
  vol_max <- max(summary_vol$volume_weighted, na.rm = TRUE)
  vol_mid <- mean(c(vol_min, vol_max))

  moci_max <- max(abs(summary_moci$moci), na.rm = TRUE)
  scale_factor <- (vol_max - vol_min) / (2 * moci_max)

  ggplot() +

    # Species lines (viridis)
    geom_line(data = summary_vol,
              aes(x = factor(year),
                  y = volume_weighted,
                  color = species_code,
                  group = species_code),
              linewidth = 1) +

    geom_point(data = summary_vol,
               aes(x = factor(year),
                   y = volume_weighted,
                   color = species_code),
               size = 3) +

    # MOCI (shared, dashed)
    geom_line(data = summary_moci,
              aes(x = factor(year),
                  y = (moci * scale_factor) + vol_mid,
                  group = 1),
              color = "grey30",
              linewidth = 1.2,
              linetype = "dashed") +

    scale_y_continuous(
      name = "Weighted mean body volume (mm3)",
      sec.axis = sec_axis(~ (. - vol_mid) / scale_factor,
                          name = "MOCI")) +

    scale_color_viridis_d(option = "D", end = 0.85) +

    labs(
      x = "Year",
      color = "Species",
      title = paste(title_label, "-", sex_filter)) +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r}
plot_adultf_species_compare(df_vol)
```

scatter plot?

```{r}
ggplot(ndif_df, aes(x = moci, y = volume_weighted, color = type)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d() +
  labs(
    x = "MOCI",
    y = "Weighted mean body volume (mm3)",
    color = "Type",
    title = "NDIF"
  ) +
  theme_minimal()
```

# Map

This part needs a lot of troubleshooting and editing: - location data is not complete (only 2005 to 2021, 3 years are missing) - Map needs bathymetry layer and fixed transect lines - Problem points/jars: line 2: 2141, 2142, line 4: 2131, 2134, 2137, 2140

```{r}
df_map <- df %>%
  left_join(
    loc_data %>%
      select(ID_Jar, Lon_IN, Lat_IN),
    by = c("jar" = "ID_Jar"))
```

making a clean df:

```{r}
df_map_clean <- df_map %>%
  select(jar, cruise, Lat_IN, Lon_IN, base_date, line, station) %>%
  distinct() %>%
  filter(!is.na(Lat_IN), !is.na(Lon_IN))
```

creating a map: (add fixed station points, check points that show land coordinates)

```{r}

pal <- colorFactor(viridis(3, option = "D"), df_map_clean$line)


leaflet(df_map_clean) %>%
  addTiles() %>%  # base map
  addCircleMarkers(
    lng = ~Lon_IN,
    lat = ~Lat_IN,
    radius = 5,
    stroke = FALSE,
    fillOpacity = 0.8,
    color = ~pal(line),  # color by line
    label = ~paste("Jar:", jar, "<br>Cruise:", cruise, "<br>Line:", line)) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~line,  # must match palette
    title = "Line",
    opacity = 1
  )
```
