---
title: "Krill Size"
subtitle: "Changes in krill size in CCE"
author: Lala Movsesian
format: html
---

# Setup

loading packages and data

```{r}
library(here)
library(readxl)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest) 

raw_data <- read_excel(here::here("data", "raw", "krill_size.xlsx"))
```

# Tidying data

### Date

Extracting year and month information from the cruise column and converting it to a convenient format. Creating base_date in the yyyy-mm-01 format to make sure it is treated as date in R. Used "cruise" column as reference rather than "date", since some sampling is done early in the next month (e.g. sampling date is 7/1 but it's from june cruise)

```{r}
#create column with year and month
df <- raw_data %>%

  mutate(base_date = make_date(
      year  = 2000 + as.numeric(substr(cruise, 4, 5)),
      month = as.numeric(substr(cruise, 6, 7)),
      day   = 1))%>% 
 #move column 
  relocate(base_date, .after = date)

```

### Life Stage

Changing all imm m to m, and imm f to f for consistency, since some samples include rows labeled as m in lower than 11mm ranges.

```{r}
df <- df %>%
  mutate(
    type = case_when(
      type == "imm m" ~ "m",
      type == "imm f" ~ "f",
      TRUE ~ type))
```

### Species

Filtering data: only including species of interest, only including juv, f and m

```{r}
df <- df %>%
  filter(species_name %in% c(
    "EUP Euphausia pacifica",
    "EUP Nematoscelis difficilis",
    "EUP Thysanoessa spinifera")) %>%
  
  # Keep only relevant types
  filter(type %in% c("juv", "f", "m"))
```

Adding code names for convenience

```{r}
df <- df %>%
  mutate(
    species_code = case_when(
      species_name == "EUP Euphausia pacifica" ~ "EPAC",
      species_name == "EUP Thysanoessa spinifera" ~ "TSPIN",
      species_name == "EUP Nematoscelis difficilis" ~ "NDIF")) %>%
  #move column
relocate(species_code, .after = species_name)
```

# Temporal changes in krill size

Testing interannual variability using Linear Mixed Models because:

-   nested data (no need to collapse data to means)

-   not independent data (avoiding pseudoreplication and inflated significance)

-   considering uneven sampling effort

-   more suitable for different months and different numbers of cruises in year

In this model, year is used as fixed effect and cruise/jar as random factors, therefore testing if variability among years is larger than within-jar/cruise.

downsides: Some years have more cruises/samples than others, which can influence the precision Years are not considered directional, and this model will not show overall trends. Considering that the main goal is to identify changes in response to climate anomalies, this is not a problem.

## uneven sampling effort

checking number of jars per cruise:

```{r}
jars_per_cruise <- df %>%
  group_by(base_date) %>%
  summarise(n_jars = n_distinct(jar))
```

checking number of jars per line per cruise

```{r}
jars_per_cruise_line <- df %>%
  group_by(base_date, line) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop")
```

checking cruises with lines that have more or less than 3 samples:

```{r}
problem_cruises <- df %>%
  group_by(base_date, line) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop") %>%
  filter(n_jars != 3)
```

### Months

checking cruises per months:

```{r}
cruise_per_month <- df %>%
    group_by(month = month(base_date)) %>%
  summarise(n_cruises = n_distinct(base_date))
```

Months of sampling represent seasons and have important data regarding seasonal variability. Months with sufficient amount of data will be used as random factors.

### Lines

checking cruises that don't have all 3 lines

```{r}
#create a list of number of lines per cruise
lines_per_cruise <- df %>%
  group_by(base_date) %>%
  summarise(n_lines = n_distinct(line), .groups = "drop") %>%
#check which cruises are missing lines
  filter(n_lines < 3)

#find out which lines are missing
missing_lines <- df %>%
  distinct(base_date, line) %>%  # which lines exist
  complete(base_date, line = c("2", "4", "6")) %>%  # create all expected combos
  anti_join(df %>% distinct(base_date, line), by = c("base_date", "line"))
```

### Station

checking stations of each line across the years, revealing multiple station per line, some N/A, some have ".", some "TT". conclusion: do not use station column as random factors

```{r}
station_per_line <- df %>%
  group_by(line, station) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop") 
```

### Diel

estimating how many night cruises by listing cruises with multiple tows (\>3 samples per line), don't have time data yet. Since number of night samples is way lower than day, night samples will be excluded from the model.

```{r}
night_cruises <- df %>%
  group_by(base_date, line) %>%
  summarise(n_jars = n_distinct(jar), .groups = "drop") %>%
  filter(n_jars > 3)
```

Using mid point size per row with length data only per type (juv/f/m) per jar per size range

```{r}
df_lmm <- df %>%
  mutate(length_mid = (length_l + length_h) / 2) %>%
relocate(length_mid, .after = length_h)
```

Calculating one size per type per jar

```{r}
df_lmm <- df_lmm %>%
  group_by(species_code, jar, type, base_date) %>%
  summarise(length_weighted = sum(length_mid * tot_num) / sum(tot_num),
    .groups = "drop")
```

excluding cruises from feb (n=1), apr (n=2), and oct (n=3)

```{r}
df_lmm <- df_lmm %>%
  filter(!month(base_date) %in% c(2, 4, 10))

```

## Euphausia pacifica

filtering species and N/As:

```{r}
epac_df <- df_lmm %>%
  filter(species_code == "EPAC") %>%
  filter(!is.na(length_weighted))
```

Creating separate dataframe for adults and juveniles:

```{r}
epac_adult <- epac_df %>%
  filter(type %in% c("f", "m"))

epac_juv <- epac_df %>%
  filter(type == "juv")
```

figure showing the size of adults over the years:

```{r}
ggplot(epac_adult, aes(x = factor(year(base_date)), y = length_weighted, color = type)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  scale_color_manual(values = c("f" = "#E64B35", "m" = "#4DBBD5")) +

  labs(x = "Year", y = "Weighted mean body length (mm)", color = "Type") +

  theme_minimal(base_size = 13) +

  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

figure showing the size of juveniles over the years:

```{r}
ggplot(epac_juv, aes(x = factor(year(base_date)), y = length_weighted)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  labs(x = "Year", y = "Weighted mean body length (mm)") +

  theme_minimal(base_size = 13) +
theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

Running model with epac adults: first checking variables

```{r}
epac_adult <- epac_adult %>%
  mutate(year = factor(year(base_date)),
    cruise = factor(base_date),
    jar = factor(jar),
    type = factor(type))
```

running mixed models:

```{r}
model_epac <- lmer(length_weighted ~ year + type + (1 | cruise/jar),
  data = epac_adult)

summary(model_epac)
```

running anova

```{r}
anova(model_epac)
```

making a plot for years:

```{r}
library(emmeans)
year_means <- emmeans(model_epac, ~ year)
plot(year_means)
```

## Thysanoessa spinifera

filtering species and N/As:

```{r}
tspin_df <- df_lmm %>%
  filter(species_code == "TSPIN") %>%
  filter(!is.na(length_weighted))
```

Creating separate dataframe for adults and juveniles:

```{r}
tspin_adult <- tspin_df %>%
  filter(type %in% c("f", "m"))

tspin_juv <- tspin_df %>%
  filter(type == "juv")
```

figure showing the size of adults over the years:

```{r}
ggplot(tspin_adult, aes(x = factor(year(base_date)), y = length_weighted, color = type)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  scale_color_manual(values = c("f" = "#E64B35", "m" = "#4DBBD5")) +

  labs(x = "Year", y = "Weighted mean body length (mm)", color = "Type") +

  theme_minimal(base_size = 13) +

  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

figure showing the size of juveniles over the years:

```{r}
ggplot(tspin_juv, aes(x = factor(year(base_date)), y = length_weighted)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  labs(x = "Year", y = "Weighted mean body length (mm)") +

  theme_minimal(base_size = 13) +
theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

## Nematoscelis difficilis

filtering species and N/As:

```{r}
ndif_df <- df_lmm %>%
  filter(species_code == "NDIF") %>%
  filter(!is.na(length_weighted))
```

Creating separate dataframe for adults and juveniles:

```{r}
ndif_adult <- ndif_df %>%
  filter(type %in% c("f", "m"))

ndif_juv <- ndif_df %>%
  filter(type == "juv")
```

figure showing the size of adults over the years:

```{r}
ggplot(ndif_adult, aes(x = factor(year(base_date)), y = length_weighted, color = type)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  scale_color_manual(values = c("f" = "#E64B35", "m" = "#4DBBD5")) +

  labs(x = "Year", y = "Weighted mean body length (mm)", color = "Type") +

  theme_minimal(base_size = 13) +

  theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

figure showing the size of juveniles over the years:

```{r}
ggplot(ndif_juv, aes(x = factor(year(base_date)), y = length_weighted)) +

  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +

  stat_summary(fun = mean, geom = "line", aes(group = type), linewidth = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +

  labs(x = "Year", y = "Weighted mean body length (mm)") +

  theme_minimal(base_size = 13) +
theme(panel.grid.minor = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```
